Exploitation Techniques and Mitigations

### 2016-01-28 docs/paper.tex
Paper: Fix errors and reword some parts

We highly recommend checking them out and having a look at their material for
further details apart from the given references.

@@

Exploiting binaries was comparatively easy ten to fifteen years ago. There were
no special mitigation mechanisms in place denying even the most simplest
exploits. This is the point in time where we will start off. First we talk
about two very simple exploits, namely the Format String Exploit and the Buffer
Overflow in combination with Shell Code. Although there is a huge collection of
exploitation techniques known to the public, we will only look at a very small
fraction of them in this project.

@@

After that, both techniques are introduced, followed by the first mitigation
technique, Data Execution Prevention (DEP). From there on we will keep on using
the buffer overflow technique with some adaptations to circumvent DEP. At that
point Return Oriented Programming (ROP) is introduced, directly leads to
Address Space Layout Randomization (ASLR) the follow-up mitigation mechanism.
Again the buffer overflow technique can be adapted to break ASLR through the
use of additional information.

Since neither DEP nor ASLR provide significant protection against even this
simple technique, an additional mitigation has been put into place in the form
of Stack Cookies.

An outlook will be given after bypassing Stack Cookies by looking at Control
Flow Integrity (CFI).

Examples will be provided along the way support the reader and provide some
additional explanation. Finally we will conclude with a word about other
architectures (x86\_64 and ARM) followed by a recap about this project.

@@

Throughout this work we assume that we know the target binary (and the
libraries it uses). Let us show that this assumption is quite reasonable to
make by looking through the eyes of the adversary. An attacker who wants to
penetrate a target machine would most likely choose the easiest path, by
exploiting the weakest link. Most machines relevant to an attacker's interest
will provide multiple services. Consider following scenario:

The main server of a small business company runs a homemade communication
service for interaction between them and their clients. The attacker has no
access to the source or binary of this communication service's daemon running
on the target machine. But along with it a commonly used web server is
listening on port 80. Getting the source (and binary) of the web server is much
easier therefore an attacker would pick this entry point over the communication
service daemon.

@@

\Cref{src:http_response} shows a possible response of a web server when
receiving an invalid request. The web server tells us his exact version and
since it also provides information about the operating system (distribution) an
attacker can easily mimic this setup to test and tweak his exploits. Exploits
may already be known to the public if the used version is not up-to-date. An
attacker could easily reuse them.

@@

At the time these techniques (and their related mitigations) were established,
x86 was the most common platform. The majority of related material found on the
internet covers x86, and many exploitation techniques can be translated from
x86 to other architectures with ease.

A more detailed overview can be found on
Wikipedia\footnote{\url{https://en.wikipedia.org/wiki/X86}} and if this is not
enough for you, consider the Intel
Manual\footnote{\url{https://www-ssl.intel.com/content/www/us/en/processors/architectures-software-developer-manuals.html}}
for a more profound insight.

@@

\Cref{fig:registers} (from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=X86&oldid=696308590\#/media/File:Table_of_x86_Registers_svg.svg}})
shows an overview of registers available on the x86 platform. While there are
dedicated registers for floating pointer operations and registers with hardware
protection (segment registers) we will only focus on nine commonly used
registers.

@@

The instruction pointer \texttt{EIP} points to the next instruction in memory
which will be executed the subsequent cycle. Stack pointer \texttt{ESP} and
base pointer \texttt{EBP} are used for stack management which is vital to call
and return from multiple functions properly. The remaining six registers are
used for arithmetic and memory operations as well as passing arguments
(parameters) for system calls. Their values can either be interpreted as
integers or pointers.

Note that these registers can be addressed partially allowing one to write only
to the lower \SI{16}{bit}, for example, as displayed in
\cref{fig:single_register} taken from \textit{null
programm}\footnote{\url{http://nullprogram.com/img/x86/register.png} on
December 2015}.

The CPU comes with protection mechanisms which allows the operating system
kernel to limit other processes' privileges. This mechanism is known as
\textit{protection rings} (Ring 0 -- Ring 3). The kernel runs \emph{in} Ring 0
(most privileged) and switches to Ring 3 (least privileged) when a normal
process is scheduled. A system call is invoked by the process if it needs
something beyond its scope. The kernel takes over, deals with the request and
returns execution back to the process. This is known as \textit{context switch}
and switching between Rings happens along with it.

@@

As already mentioned in the previous paragraph, a process only has limited
capabilities and the kernel has to take over to fulfill certain (more
privileged) operations. The operating system's documentation tells you which
system calls are available (on which platform) and what parameters
each of them requires. Let us illustrate this with an example: On x86 Linux the
system call
number 4 (starting from 0) is the \texttt{sys\_write} system call which writes
data to a file descriptor. It takes three arguments, the file descriptor to
write to, a pointer to the start of the data which should be written and the
length of the data. The number of the system call together with these three
parameters are placed in the \texttt{EAX}, \texttt{EBX}, \texttt{ECX},
\texttt{EDX} respectively. To invoke the system call issue following
instruction:

@@

Nowadays you may encounter a different mechanism for system calls using Virtual
Dynamic Shared Objects (vDSO). This goes beyond our scope here, we will use the
previously mentioned mechanism in our exploits as they work side by side.
Consult the corresponding man
page\footnote{\url{http://man7.org/linux/man-pages/man7/vdso.7.html}} for
further reading.

@@

Physical memory is managed by the operation system kernel by utilising the
Memory Management Unit (MMU). Each process' address space is virtualized and
memory operations are translated on-the-fly by the MMU. Physical memory is
segmented into \textit{pages} (typically \SI{4}{\kibi\byte} in size) and each
page can be mapped \emph{into} the virtual address space of one or more
(shared) processes.~\cite[pp.~400]{unix_interals}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.35\textwidth]{gfx/little_endian.pdf}
    \caption{Placement of bytes in memory in little-endian}
    \label{fig:little_endian}
\end{figure}

Endianness refers to the byte order used when storing data in memory (or
transmitting it over the network). x86 uses little-endian which is described in
\cref{fig:little_endian} (from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=Endianness&oldid=696417697\#/media/File:Little-Endian.svg}}).
The least significant byte of a word is placed at the lower memory address and
successive bytes are placed as the memory address increases. We will later
refer back to this when needed. The related Wikipedia
page\footnote{\url{https://en.wikipedia.org/wiki/Endianness}} goes into more
detail about this than we need.

@@

Taking a closer look at \texttt{printf} we can see that its first argument is a
format string followed by a variable number of additional arguments.
\texttt{stdarg.h} describes a common implementation of this together with a
small example in their man
page\footnote{\url{http://linux.die.net/man/3/stdarg}}. As in that example,
\texttt{printf} trusts you that the number of arguments supplied is equal (or
greater) than the number of formatters. Calling \texttt{printf} with the format
string \texttt{"\%d + \%d = \%d"}, for instance, assumes that (at least) three
additional arguments are given.

@@

            > echo '%08x' | ./main
            You entered:
            bfd98ed4
        \end{code*}
    \end{minipage}
    \caption{Program vulnerable to Format String Exploits}
    \label{src:format_string_exploit}
\end{listing}

The exploit comes from the notion that a format string provided by an attacker
gets interpreted. The program shown in \cref{src:format_string_exploit} will
take an arbitrary string from \texttt{stdin} and pass it on to \texttt{printf}.
For simple inputs (not containing formatter) this works fine. But as soon as
formatter are provided, \texttt{printf} accesses the locations where the
corresponding arguments \emph{would} be located. From the calling convention
described in \cref{sub:calling_convention} we know that these arguments
\emph{would} be located on the stack, therefore \texttt{printf} will print
whatever lies on the stack at these positions instead.

@@

We would like to point the reader to the book \textit{Hacking: The Art of
Exploitation}~\cite[pp.~167]{art_of_exploitation} for more details about this
and following techniques. We will come back to this technique later on to show
that \texttt{printf} enables even more sophisticated attacks.

@@

The second type of exploits we'll look at is known as Buffer Overflows and as
on may already derive from the name, this is about submitting more data to a
buffer than it was originally designed for. This setup can be exploited when
bound checking is done wrong or not at all. An attacker is therefore able to
overwrite memory behind the buffer's location.

@@

The three mitigation mechanisms DEP, ASLR and Stack Cookies are enabled by
default nowadays and as mentioned in the introduction we start off at a pointer
where these mechanisms were not. So to run the provided examples we first have
to disable them. DEP and Stack Cookies can be disabled via compiler flags to
the extend necessary using \texttt{-z execstack} and
\texttt{-fno-stack-protector} respectively.

ASLR can be disabled globally so that \emph{new} process' has an unscrambled
memory layout:

@@

Writing \texttt{2} instead of \texttt{0} will switch ASLR back to its default
state. Root privileges are, of course, required for this. There is also another
way by invoking \texttt{setarch}:

@@

The consequences of an exploited buffer overflow depend on where the buffer is
located. The most interesting location would of course be the stack because,
apart from local variables and arguments, it holds the return address of a
function. But buffers located inside the heap or static may also be viable
options. Common terms related to these scenarios are \textit{stack smashing}
and \textit{heap corruption}. We will talk about heap corruption later on when
breaking ASLR, for now we focus our attention on stack smashing.

@@

Let's start off by examining the stack containing a buffer \texttt{c} as local
variable, see \cref{fig:stack_frame}. Right now the buffer holds the string
\texttt{"hello"} followed by a terminator. Since it has been allocated
to hold a maximum of \SI{12}{\byte} this fits. If data is written to the buffer
larger than \SI{12}{\byte} the following variable (or parameter) \texttt{bar}
will be overwritten, followed by the saved frame pointer and the return
address. If even more data is supplied the following stack frame will be
overwritten in the same manner.

If an attacker can provide the data written to the buffer and no (or wrong)
bound checking is done, he is able to inject arbitrary (malicious) code into
the stack frame. This could be, for instance, be used to overwrite a flag
indication whether an authentication has been performed successfully or not.
But since this is pretty forward let's go beyond that and see what happens when
changing the return address.

@@

As shown in \cref{src:buffer_overflow} we have a buffer suited for
\SI{20}{\byte} but without any bound checking. If the provided input is longer,
we will be able to overwrite the return address. Let's have a look at the
resulting binary utilizing \texttt{objdump}.

@@

\texttt{mordor} has been executed successfully. Despite the segmentation
fault one can see that return address has been overwritten successfully.

@@

While this is neat and can certainly be useful to an adversary, stack smashing
also enables us to inject arbitrary code into a program. Contrary to the
previous section the target machine will execute code provided by the attacker.
This can be achieved by bending the return address into the buffer used for the
exploit. Provided instructions will be executed upon return. Shell code is a
piece of (binary) code which opens up a shell that reads and executes commands
from an attacker. This example is taken from Dhaval Kapil's
blog\footnote{\url{https://dhavalkapil.com/blogs/Shellcode-Injection/} on
December 2015} there is also a section about this in \textit{Hacking: The Art
of Exploitation}~\cite[pp.~281]{art_of_exploitation}. There is also a
comprehensive article \cite{phrack_stack_smash} about Stack Smashing on
\texttt{phrack}\footnote{\url{http://phrack.org/}}.

@@

The piece of assembly shown in \cref{src:shell_code} sets up the parameters for
the \texttt{execve} system call and than invokes it to replace the currently
running process with a shell. \texttt{execve} takes three arguments, a string
of the program to execute (here \texttt{"/bin//sh"} + terminator), a list of
arguments for that program and a list of environment variables. Its system call
number is 11 and it will accept \texttt{NULL} for both lists. The double slash
in the first argument is used to prevent null bytes inside the shell code. The
function which reads the shell code may truncate it upon reading a null byte,
therefore we have to work around this without changing the underlying
semantics.

@@

Finding the starting location of our buffer will be a little bit more
complicated, we cannot read it directly from the binary of the target program
so, in \cref{src:shell_code2}, we'll examine it in a debugger .

@@

Now we know that the buffer will be located at \texttt{0xbffff4bc} (saved base
pointer will be at \texttt{0xbffff528}) at runtime, but it may be offset a few
bytes when run without debugger. This happens because environment variables and
meta information, like the program name, determine the stack starting position
(stack is placed right before environment variables). Hence we may not hit the
first instruction of our shell code right away, but since the buffer is bigger
than the actual payload we can improve our odds by prefixing the shell code
with \texttt{NOP} instructions. As long as the return address points somewhere
into this sequence of \texttt{NOP}s the CPU will \emph{slide} to the first
instruction of the shell code. Therefore this is known as a
\textit{\texttt{NOP} Sled}. We append some arbitrary data to the shell code as
offset to overwrite the return address. This is also illustrated in
\cref{fig:shell_code} where \textit{target} is new return address supplied by
the attacker. Using the maximum amount of \texttt{NOP}s possible would also be
a viable option, here we just went with the original example.

@@

DEP makes injection arbitrary code into binary much harder, but we still
control the return address --- so let's use it.

@@

The final \texttt{ret} statement of a gadgets is required for chaining them
together. We not only use the Buffer Overflow to control \emph{one}, but
\emph{multiple} return addresses. Because of this our payload will be made up
by a list of return addresses (the start of each gadgets) interleaved by some
padding. Each provided address will be consumed one by one at the end of each
gadgets to \emph{jump} to the next one. The payload may still contain data like
\texttt{"/bin//sh"} if necessary.

@@

The target program is displayed in \cref{src:rop_example}. We will not be able
to inject and execute shell code, and there is no function present which
directly opens up a shell for us. But there are parts which can be used to do
so.

On the right hand side we see the execution of the exploit. First note that we
no longer compile the binary with \texttt{-z execstack}. We read the locations
of \texttt{not\_used} and \texttt{system} via \texttt{gdb} and note down the
corresponding addresses. \texttt{objdump} is used to have a quick glance at the
generated binary code for \texttt{vulnerable\_function} and note down the
distance between the saved base pointer the start of the buffer too (line 17).

Putting this together yields following payload: Starting with some
\texttt{'A'}s to fill the buffer followed by 4 \texttt{'B'}s to overwrite the
saved base pointer. The next part is new, we attach the address of
\texttt{system} followed by some padding and a pointer to \texttt{not\_used}.

@@

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as a Position Independent Executable (PIE). Since this is
actually the default of \texttt{gcc}, most programs' \texttt{.text} segment
will always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, so ASLR would be able to
randomize these segments too.

Breaking ASLR, even when the code is compiled with \texttt{-pie}, is easier
than it seems at first. Relocation only happens to a section at whole,
functions inside a section still share the same relative distance as they would
without ASLR. But before exploit this fact have a look at the randomized
addresses again.

Only three nybble ($3 \times \SI{4}{\bit}$) differ between multiple runs giving
us $2^{12} = 4096$ possibilities. If the scenario allows it, brute forcing is a
viable option here. But note that this changes drastically for \SI{64}{\bit}.
But we won't hassle with brute force, a better option has already been teased.

@@

Manipulate the target file used in the ROP example to print the address of
\texttt{printf} first, \emph{then} read in the payload via \texttt{stdin}. This
way you can first simulate a leaked pointer, adapt the ROP chain and run it.
Double check the distances between functions in the libraries, they may differ
with the ones used in this writeup.

@@

    \item[Random] A random canary is chosen at program start, stored
        \textit{somewhere save} and pushed into the stack upon function calls.
        The canary on the stack is checked against the one stored before
        executing the return instruction. The program is terminated upon
        mismatch. With this setup an attacker has to know the canary in order
        to overwrite the return address. Since it is chosen at random during
        program start, an attacker cannot reliably reproduce the same canary in
        his cloned setup.

@@

The canary itself is composed of a terminator (null) followed by a random
sequence of \SI{3}{\byte}. This sequence changes very time the program is run.
Feeding more data to the buffer and overflowing it this way yields termination
of the program. Note that \texttt{puts} is still executed, the termination
happens just before the return of \texttt{fun}.

@@

\subsection{Server Worker Paradigm}

Of course there are multiple paths available when trying to break the
StackGuard mechanism, as mentioned in \cite{phrack_stack_guard}. We will now
have a look at the common server worker paradigm. \Cref{src:server_worker}
shows how that paradigm looks like from a task monitors view. The server /
daemon (here \texttt{apache2}) is started with root privileges in order to
listen on a \textit{privileged} port. After the initialization has been
compiled the server forks itself multiple times to create a set of workers. In
this example the workers drop their root privileges right away by changing
their current user to \texttt{www-data}. But our focus is not on the privileges
but the problem introduced by fork with respect to the StackGuard.

@@

Again the name already tells you what this is about, a directed graph that
reflects the control-flow of a program. Different definitions exist regarding
what is actually used to compose a node and anchor them together.In our case we
will create a node out of each function and connect them at function calls.

@@

\Cref{fig:cfg} shows an example of such a graph. From this graph you can tell
that \texttt{funB} will be called from \texttt{funA} and \texttt{funE} but not,
for instance, from \texttt{funD}. This is all we care about right now. Note
that contrary to some definitions this graph is not acyclic since we also model
direct and indirect recursion as can be seen by $\texttt{funC} \to
\texttt{funC}$ and $\texttt{funcB} \to \texttt{funcD} \to \texttt{funcE} \to
\texttt{funcB}$ respectively.

@@

CFI is a big topic and, like other topics already mentioned, goes beyond the
scope of this writeup. The first pointer we will hand the reader aims at the
corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we recommend checking out the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).

>>>>

We highly recommend checking them out and having a look at their material for
further details, apart from the given references.

@@

Exploiting binaries was comparatively easy ten to fifteen years ago. There were
no special mitigation mechanisms in place denying even the most simplest
exploits. This is the point in time where we will start off. First we talk
about two very simple exploits, namely the format string exploit and the buffer
overflow in combination with shell code. Although there is a huge collection of
exploitation techniques known to the public, we will only look at a very small
fraction of them in this project.

@@

After that, both techniques are introduced, followed by the first mitigation
technique, Data Execution Prevention (DEP). From there on we will keep on using
the buffer overflow technique with some adaptations to circumvent DEP. At that
point Return Oriented Programming (ROP) is introduced, which directly leads us
to Address Space Layout Randomization (ASLR), the follow-up mitigation
mechanism. Again, the buffer overflow can be adapted to break ASLR through the
use of additional information (info leak).

Since neither DEP nor ASLR provide significant protection against even this
simple technique, an additional mitigation has been put into place in the form
of stack cookies (StackGuard).

An outlook will be given after bypassing StackGuard by looking at Control-Flow
Integrity (CFI).

Examples will be provided along the way to support the reader and provide
additional explanations. Finally we will conclude with a word about other
architectures (x86\_64 and ARM) followed by a short recap about what has been
taught in this writeup.

@@

Throughout this work we assume that we know the target binary (and the used
libraries). Let us show that this assumption is quite reasonable to make by
looking through the eyes of the adversary. An attacker who wants to penetrate a
target machine would most likely choose the easiest path --- exploiting the
weakest link. Most machines relevant to an attacker's interest will provide
multiple services. Consider following scenario:

The main server of a small business company runs a homemade communication
service for interaction between them and their clients. The attacker has no
access to the source or binary of this communication service's daemon running
on the target machine. But, along with it, a commonly used web server is
listening on port 80. Getting the source (and binary) of the web server is much
easier, therefore an attacker would pick this entry point over the
communication service daemon.

@@

\Cref{src:http_response} shows a possible response of a web server when
receiving an invalid request. The web server tells us his exact version and
since it also provides information about the operating system (distribution) an
attacker can easily mimic this setup to test and tweak his exploits. Exploits
may already be known to the public if the used version is not up-to-date. An
attacker can use, modify and build upon them.

@@

At the time these techniques (and their related mitigations) were established,
x86 was the most common computing platform. The majority of related material
found on the internet covers x86, and many techniques can be translated from
x86 to other architectures with ease.

A more detailed overview can be found on
Wikipedia\footnote{\url{https://en.wikipedia.org/wiki/X86}} and --- if this is
not enough --- consider the Intel
Manual\footnote{\url{https://www-ssl.intel.com/content/www/us/en/processors/architectures-software-developer-manuals.html}}
for a more profound insight.

@@

\Cref{fig:registers} (taken from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=X86&oldid=696308590\#/media/File:Table_of_x86_Registers_svg.svg}})
shows an overview of registers available on the x86 platform. While there are
dedicated registers for floating-point operations and registers with hardware
protection (segment registers) we will only focus on nine commonly used
registers.

@@

The instruction pointer \texttt{EIP} points to the next instruction in memory
which will be executed on the subsequent cycle. Stack pointer \texttt{ESP} and
base pointer \texttt{EBP} are used for stack management which is vital to call
and return from multiple functions properly. The remaining six registers are
used for arithmetic and memory operations as well as passing arguments
(parameters) for system calls. Their values can either be interpreted as
integer or pointer.

Note that these registers can be addressed partially, allowing one to write
only to the lower \SI{16}{\bit}, for example, as displayed in
\cref{fig:single_register} (taken from \textit{null
programm}\footnote{\url{http://nullprogram.com/img/x86/register.png} on
December 2015}).

The CPU comes with protection mechanisms which allows the operating system's
kernel to limit other processes' privileges. This mechanism is known as
\textit{protection rings} (Ring 0 -- Ring 3). The kernel runs \emph{in} Ring 0
(most privileged) and switches to Ring 3 (least privileged) when a normal
process is scheduled. A system call is invoked by the process if it needs
anything beyond its scope. The kernel takes over, deals with the request and
returns execution back to the process. This is known as \textit{context switch}
and switching between Rings happens along with it.

@@

As already mentioned in the previous paragraph, a process only holds limited
capabilities and the kernel has to take over to fulfill certain (more
privileged) operations. The operating system's documentation tells which system
calls are available (on which platform) and what parameters each of them
requires. Let us illustrate this with an example: On x86 Linux the system call
number 4 (starting from 0) is the \texttt{sys\_write} system call which writes
data to a file descriptor. It takes three arguments, the file descriptor to
write to, a pointer to the start of the data which should be written and the
length of the data. The number of the system call, together with these three
parameters are placed in the \texttt{EAX}, \texttt{EBX}, \texttt{ECX},
\texttt{EDX} registers respectively. Following instruction is issued to invoke
the system call:

@@

Nowadays you may encounter a different mechanism used for system calls,
utilizing Virtual Dynamic Shared Objects (vDSO). This goes beyond our scope
here and we will use the previously mentioned mechanism in the following
exploits as they work side by side. Consult the corresponding man
page\footnote{\url{http://man7.org/linux/man-pages/man7/vdso.7.html}} for
further reading.

@@

Physical memory is managed by the kernel through the use of a Memory Management
Unit (MMU). Each process' address space is virtualized and memory operations
are translated on-the-fly by the MMU. Physical memory is segmented into
\textit{pages} (typically \SI{4}{\kibi\byte} in size) and each page can be
\emph{mapped into} the virtual address space of one or more (shared page)
processes.~\cite[pp.~400]{unix_interals}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.35\textwidth]{gfx/little_endian.pdf}
    \caption{Byte order in little-endian}
    \label{fig:little_endian}
\end{figure}

Endianness refers to the byte order used when storing data in memory (or
transmitting it over the network). x86 uses little-endian which is described in
\cref{fig:little_endian} (taken from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=Endianness&oldid=696417697\#/media/File:Little-Endian.svg}}).
The least significant byte of a word is placed at the lower memory address and
successive bytes are placed as the memory address increases. The related
Wikipedia page\footnote{\url{https://en.wikipedia.org/wiki/Endianness}} goes
into more detail about this --- more than currently needed. We will later refer
back to this when swapping bytes because of endianness.

@@

Taking a closer look at \texttt{printf} we can see that its first argument is a
format string followed by a variable number of additional arguments. A common
implementation, together with a small example, of this is described in the man
page\footnote{\url{http://linux.die.net/man/3/stdarg}} of \texttt{stdarg.h}.
\texttt{printf} trusts the programmer that the number of arguments supplied is
equal (or greater) than the number of formatters. Calling \texttt{printf} with
the format string \texttt{"\%d + \%d = \%d"}, for instance, assumes that (at
least) three additional arguments are given.

@@

            > echo '%08x' | ./main
            You entered:
            bfd98ed4
        \end{code*}
    \end{minipage}
    \caption{Program vulnerable to format string exploits}
    \label{src:format_string_exploit}
\end{listing}

The exploit comes from the notion that a format string provided by an attacker
gets interpreted. The program shown in \cref{src:format_string_exploit} will
take an arbitrary string from \texttt{stdin} and pass it on to \texttt{printf}.
For simple inputs (not containing formatters) this works fine. But as soon as
formatters are provided, \texttt{printf} accesses the locations where the
corresponding arguments \emph{would} be located. From the calling convention
described in \cref{sub:calling_convention} we know that these arguments
\emph{would} be located on the stack, therefore \texttt{printf} will print
whatever lies on the stack at these positions instead.

@@

We would like to point the reader to the book \textit{Hacking: The Art of
Exploitation}~\cite[pp.~167]{art_of_exploitation} for more details about this
and similar techniques. We will come back to this technique later on to show
that \texttt{printf} enables even more sophisticated attacks (see
\cref{sub:printbf}).

@@

The second type of exploits we'll look at is known as buffer overflows and as
on may already derive from the name, this is about submitting more data to a
buffer than it was originally designed for. This setup can be exploited when
bound checking is done wrong or not at all. An attacker is therefore able to
overwrite memory behind the buffer's location.

@@

The three mitigation mechanisms DEP, ASLR and StackGuard are enabled by default
nowadays, but, as mentioned in the introduction, we start off at a point where
these mechanisms were not yet in place. So to run the provided examples we
first have to disable them. DEP and StackGuard can be disabled via compiler
flags to the extend necessary using \texttt{-z execstack} and
\texttt{-fno-stack-protector} respectively.

ASLR can be disabled globally so that \emph{new} processes have an unscrambled
memory layout:

@@

Writing \texttt{2} instead of \texttt{0} will switch ASLR back to its default
state. Root privileges are, of course, required for this. There is also another
way by using \texttt{setarch} to run a binary:

@@

The consequences of an exploited buffer overflow depend on where the buffer is
located. The most interesting location would of course be the stack because,
apart from local variables and arguments, it holds the return address of a
function. But buffers located inside the heap or static may also be viable
options. Common terms related to these scenarios are \textit{stack smashing}
and \textit{heap corruption}. For now we focus our attention on stack smashing.

@@

Let's start off by examining the stack containing a buffer \texttt{c} as local
variable, see \cref{fig:stack_frame}. Right now the buffer holds the string
\texttt{"hello"} followed by a terminator. Since it has been allocated to hold
a maximum of \SI{12}{\byte} this fits. If data larger than \SI{12}{\byte} is
written to the buffer, the following variable (or parameter) \texttt{bar} will
be overwritten, followed by the saved frame pointer and the return address. If
even more data is supplied the adjacent stack frame will be overwritten in the
same manner.

If an attacker can provide the data written to the buffer and no (or wrong)
bound checking is done, he is able to inject arbitrary (malicious) code into
the stack frame. This could, for instance, be used to overwrite a flag
indication whether an authentication has been performed successfully or not.
But since this is pretty forward let's go beyond that and see what happens when
changing the return address.

@@

As shown in \cref{src:buffer_overflow} we have a buffer suited for
\SI{20}{\byte} but without any bound checking. If the provided input is longer,
it will overwrite the return address. Let's have a look at the resulting binary
utilizing \texttt{objdump}.

@@

\texttt{mordor} has been executed successfully. Despite the segmentation fault
one can see that the return address has been overwritten successfully.

@@

While this is neat and can certainly be useful to an adversary, stack smashing
also enables us to inject arbitrary code into a program. Contrary to the
previous section the target machine will execute code provided by the attacker.
This can be achieved by bending the return address into the buffer used for the
exploit. Provided instructions will be executed upon return. Shell code is a
piece of (binary) code which opens up a shell that reads and executes commands
from an attacker. This example is taken from Dhaval Kapil's
blog\footnote{\url{https://dhavalkapil.com/blogs/Shellcode-Injection/} on
December 2015} there is also a section about this in \textit{Hacking: The Art
of Exploitation}~\cite[pp.~281]{art_of_exploitation}.

A more com comprehensive article \cite{phrack_stack_smash} about Stack Smashing
is available on \texttt{phrack}\footnote{\url{http://phrack.org/}}.

@@

The piece of assembly shown in \cref{src:shell_code} sets up the parameters for
the \texttt{execve} system call and then invokes it to replace the currently
running process with a shell. \texttt{execve} takes three arguments, a string
of the program to execute (here \texttt{"/bin//sh"} + terminator), a list of
arguments for that program and a list of environment variables. The
corresponding system call number is 11 and \texttt{NULL} will be accepted for
both lists. The double slash in the first argument is used to prevent null
bytes inside the shell code. The function which reads the shell code may
truncate it upon reading a null byte, therefore we have to work around this
without changing the underlying semantics.

@@

We'll examine the target binary in a debugger (\cref{src:shell_code2}) to find
the starting location of our buffer.

@@

Now we know that the buffer will be located at \texttt{0xbffff4bc} (saved base
pointer will be at \texttt{0xbffff528}) at runtime, but it may be offset a few
bytes when run without a debugger. This happens because environment variables
and meta information, like the program name, determine the stack starting
position (stack is placed right before environment variables). Hence we may not
hit the first instruction of our shell code right away, but since the buffer is
bigger than the actual payload we can improve our odds by prefixing the shell
code with \texttt{NOP} instructions. As long as the return address points
somewhere into this sequence of \texttt{NOP}s the CPU will \emph{slide} to the
first instruction of the shell code. Therefore this is known as a
\textit{\texttt{NOP} Sled}. We append some arbitrary data to the shell code as
offset to overwrite the return address. This is also illustrated in
\cref{fig:shell_code} where \textit{target} is the new return address supplied
by the attacker. Using the maximum amount of \texttt{NOP}s possible would also
be a viable option, here we just went with the original example.

@@

DEP makes injecting arbitrary code into binary much harder, but we still
control the return address --- so let's use it.

@@

The final \texttt{ret} statement of a gadgets is required for chaining them
together. We not only use the buffer overflow to control \emph{one}, but
\emph{multiple} return addresses. Because of this our payload will be composed
of a list of return addresses (the start of each gadgets) interleaved by some
padding. Each provided address will be consumed one by one at the end of each
gadgets to \emph{jump} to the next one. The payload may still contain data like
\texttt{"/bin//sh"} if necessary.

@@

The target program is displayed in \cref{src:rop_example}. We will not be able
to inject and execute shell code --- and there is no function present which
directly opens up a shell for us. But there are parts which can be glued
together to do so.

On the right hand side we see the execution of the exploit. First note that we
no longer compile with \texttt{-z execstack}. We read the locations of
\texttt{not\_used} and \texttt{system} via \texttt{gdb} and note down the
corresponding addresses. \texttt{objdump} is used to have a quick glance at the
generated binary code for \texttt{vulnerable\_function} and note down the
distance between the saved base pointer the start of the buffer too (line 17).

Following payload can be established with the gathered information: Starting
with some \texttt{'A'}s to fill the buffer followed by 4 \texttt{'B'}s to
overwrite the saved base pointer. The next part is new, we attach the address
of \texttt{system} followed by some padding and a pointer to
\texttt{not\_used}.

@@

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as a Position Independent Executable (PIE). Since this is
actually the default of \texttt{gcc}, most programs' \texttt{.text} segment
will always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, so ASLR would be able to
randomize these segments too, but you'd have to compile every relevant package
again instead of using the distribution vendor's pre-compiled binary.

Breaking ASLR, even when the code is compiled with \texttt{-pie}, is easier
than it seems at first. Relocation only happens to a section as a whole,
functions inside a section still share the same relative distance as they would
without ASLR. But before exploit this fact, have a look at the randomized
addresses again.

Only three nybble ($3 \times \SI{4}{\bit}$) differ between multiple runs giving
us $2^{12} = 4096$ possibilities. If the scenario allows it, brute forcing
would be a viable option here. But note that this changes drastically for
\SI{64}{\bit}. But we won't hassle with brute force, a better option has
already been teased.

@@

Manipulate the target source code used in the ROP example to print the address
of \texttt{printf} first, \emph{then} read in the payload via \texttt{stdin}.
This way you can first simulate a leaked pointer, adapt the ROP chain and run
it. Double check the distances between library functions, they may differ with
the ones used in our example.

@@

    \item[Random] A random canary is chosen at program start, stored
        \textit{somewhere save} and pushed onto the stack upon function calls.
        The canary on the stack is compared with the one stored save before
        executing the return instruction. The program is terminated on
        mismatch. With this setup an attacker has to know the canary in order
        to overwrite the return address via a buffer overflow. Since it is
        picked at random during program start, an attacker cannot reliably
        reproduce the same canary in his cloned setup.

@@

The canary itself is composed of a terminator (null) followed by a random
sequence of \SI{3}{\byte}. This sequence changes every time the program is run.
Feeding more data to the buffer and overflowing it this way yields termination
of the program. Note that \texttt{puts} is still executed, the termination
happens just before the return of \texttt{fun}.

@@

\subsection{Server-Worker Paradigm}

Of course there are multiple paths available when trying to break the
StackGuard mechanism, as mentioned in \cite{phrack_stack_guard}. We will now
have a look at the common server-worker paradigm. \Cref{src:server_worker}
shows how that paradigm looks like from a task monitor's view. The server /
daemon (here \texttt{apache2}) is started with root privileges in order to
listen on a \textit{privileged} port. After the initialization has been
compiled, the server forks itself multiple times to create a set of workers. In
this example the workers drop their root privileges right away by changing
their current user to \texttt{www-data}. But our focus is not on the privileges
but the problem introduced by fork with respect to the StackGuard.

@@

Again the name already tells you what this is about, a directed graph that
reflects the control-flow of a program. Different definitions exist regarding
what is actually used to compose a node and anchor them together. In our case
we will create a node out of each function and connect them at function calls.

@@

\Cref{fig:cfg} shows an example of such a graph. From it one can tell that
\texttt{funB} will be called from \texttt{funA} and \texttt{funE} but not, for
instance, from \texttt{funD}. This is all we care about right now. Note that
contrary to some definitions this graph is not acyclic since we also model
direct and indirect recursion as can be seen by $\texttt{funC} \to
\texttt{funC}$ and $\texttt{funcB} \to \texttt{funcD} \to \texttt{funcE} \to
\texttt{funcB}$ respectively.

@@

CFI is a big topic and, like other topics already mentioned, goes beyond the
scope of this writeup. The first pointer we hand the reader aims at the
corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we recommend checking out the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).
